The \dealii{} Library, Version 9.4

### 2022-06-23 9.4/paper.tex
Minor language fixes.

The matrix-free infrastructure of \dealii{} is used to solve problems
without assembling matrices, providing only the action of the matrix
instead. It has seen a number of new features in this release, the most
notable ones being:
\begin{itemize}
\item Improved support for computations with Hessians of shape functions: just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This could enable, for example, writing
a matrix-free version of step-47, which solves the biharmonic equation with the discontinuous Galerkin method.
\item Cell-centric loops now also allow access to gradients and Hessians
of neighboring cells on faces. The major difficulty here was the relative orientation
of cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in several
contexts. Examples are simulations with sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or non-cut cells, making vectorization of these operations potentially more complicated. Previous functionality has provided the option of masking certain cells in cell batches, which works well if
the code paths do not diverge too much. Another way is to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, \texttt{MatrixFree::reinit()}
might be too expensive if recategorization needs to happen very frequently to follow the dynamics of a system, e.g., in each time step. Despite some overhead compared to static matrix-free loops, the new feature can be the best option in these scenarios.
\item Initial support for H(div)-conforming elements with Piola transform, based on Raviart--Thomas finite element spaces with the updated class \texttt{FE\_RaviartThomasNodal}, has been added. This feature is currently limited to meshes in standard orientation and affine geometries. Full support and performance optimizations will be provided in a future release.
\item Selected matrix-free algorithms can now exploit additional data locality
  between the matrix-vector product and vector operations happening nearby in
  an algorithm. Interfaces have been added to both the
  \texttt{PreconditionChebyshev} class (a frequently used smoother in multigrid
  methods) and the conjugate gradient implementation in \texttt{SolverCG}.
  From the user's perspective, an operator needs to define a \texttt{vmult} operation,
  taking two additional \texttt{std::function} arguments. The first function
  defines the operation to be scheduled on the vector entries before the
  matrix-vector product touches them, and the second what happens
  afterwards. The new features also include a renumbering to maximize data
  locality. The theory is described in the
  contribution~\cite{kronbichler2022cg}.
\end{itemize}
Besides these new features, we improved the performance of
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on in-place interpolation and sum factorization, similar
to what was already done in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of ten.

Finally, we have performed a major restructuring of internals
of the \texttt{FEEvaluation} classes. This reduces some overheads for low polynomial degrees and will enable us to add support for new element types in the future.

@@

In release 9.3~\cite{dealII93}, we added support for global-coarsening multigrid in
addition to the established local-smoothing infrastructure. Global
coarsening algorithms smoothen on the whole computational domain on each
multigrid level, which is obtained by coarsening the finest cells of
the next finer multigrid level.
For this purpose, we use a sequence of triangulations, and we perform
the smoothing only on their active levels. To create the sequence of
triangulations, one can use the functions \texttt{MGTransferGlobalCoarseningTools::create\_geometric\_coarsening\_sequence()}. A new version takes an
instance of \texttt{RepartitioningPolicyTools::Base} (see Subsection~\ref{sec:repartitioning}) as argument, which allows to specify the parallel
distribution of each multigrid level (in contrast to the fixed first-child policy in the case of local smoothing). These features have been developed and tuned for running on supercomputer scale
with complicated coarse meshes as presented in \cite{kronbichler2021next}.
Furthermore, we added support for block vectors,
fixed a number of limitations, and performed performance optimizations of
the transfer operator; particularly, the redundant copy from/to temporary vectors
has been eliminated. Furthermore, hanging-node constraints are applied
efficiently in the same way as in the matrix-free loops (see Subsection~\ref{sec:mf}).

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \dealii{} was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order to judge the benefits
of one approach against the other, \dealii{} provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()}
in the \texttt{MGTools} namespace for the  estimation of the imbalance during, e.g.,
smoothing and the
communication efficiency during intergrid transfer purely based on the given mesh.
% MK: I would not add this part, it does not really fit into this paper as
% there is nothing to report at this point (and we should then add references
% to the actual literature).
%The publication \cite{munch2022gc} also points out that not
%all types of smoothers are applicable for global coarsening due to the
%presence of hanging nodes, which is a motivation to add new smoother types
%to \dealii in the future.

@@

Table~\ref{tab:particle_timing} shows that all particle operations are
much faster in \dealii 9.4 than in version 9.3. In particular, operations that depend strongly on particle storage structure and require few fixed computations (like iteration and sorting) benefit massively from the above mentioned optimizations. We note that the exact gains will depend strongly on the exact combination of geometry, mapping, dimensionality, and number of particles per cell in any specific model, and can be smaller or larger than the measurements provided here.

@@

Many problems in parallel computations can be stated in the following
way: Each process in a parallel universe has a number of queries to
send to other processes who do not know that they will be asked, and
who will then have to respond with replies. This problem is solved by
``consensus algorithms''~\cite{hoefler2010scalable}. An example of where this problem appears is
given in Section~\ref{sec:repartitioning}.

@@

There are various ways to reference \dealii{}. To acknowledge the use of
the current version of the library, \textbf{please reference the present
  document}. For up to date information and a bibtex entry
see
\begin{center}
  \url{https://www.dealii.org/publications.html}
\end{center}

>>>>

The matrix-free infrastructure of \dealii{} is used to solve problems
without assembling matrices, providing only the action of the matrix
instead. It has seen a number of new features in this release, the most
notable ones being:
\begin{itemize}
\item Improved support for computations with Hessians of shape functions: just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This could enable, for example, writing
a matrix-free version of step-47, which solves the biharmonic equation with the discontinuous Galerkin method.
\item Cell-centric loops now also allow access to gradients and Hessians
of neighboring cells on faces. The major difficulty here was the relative orientation
of cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in several
contexts. Examples are simulations with sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or non-cut cells, making the vectorization of these operations potentially more complicated. Previous functionality has provided the option of masking certain cells in cell batches, which works well if
the code paths do not diverge too much. Another way is to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, \texttt{MatrixFree::reinit()}
might be too expensive if recategorization needs to happen very frequently to follow the dynamics of a system, e.g., in each time step. Despite some overhead compared to static matrix-free loops, the new feature can be the best option in these scenarios.
\item Initial support for H(div)-conforming elements with Piola transform, based on Raviart--Thomas finite element spaces with the updated class \texttt{FE\_RaviartThomasNodal}, has been added. This feature is currently limited to meshes in standard orientation and affine geometries. Full support and performance optimizations will be provided in a future release.
\item Selected matrix-free algorithms can now exploit additional data locality
  between the matrix-vector product and vector operations happening nearby in
  an algorithm. Interfaces have been added to both the
  \texttt{PreconditionChebyshev} class (a frequently used smoother in multigrid
  methods) and the conjugate gradient implementation in \texttt{SolverCG}.
  From the user's perspective, an operator needs to define a \texttt{vmult} operation,
  taking two additional \texttt{std::function} arguments. The first function
  defines the operation to be scheduled on the vector entries before the
  matrix-vector product touches them, and the second what happens
  afterward. The new features also include a renumbering to maximize data
  locality. The theory is described in the
  contribution~\cite{kronbichler2022cg}.
\end{itemize}
Besides these new features, we improved the performance of
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on in-place interpolation and sum factorization, similar
to what was already done in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of ten.

Finally, we have performed a major restructuring of the internals
of the \texttt{FEEvaluation} classes. This reduces some overheads for low polynomial degrees and will enable us to add support for new element types in the future.

@@

In release 9.3~\cite{dealII93}, we added support for global-coarsening multigrid in
addition to the established local-smoothing infrastructure. Global
coarsening algorithms smoothen over the whole computational domain on each
multigrid level, which is obtained by coarsening the finest cells of
the next finer multigrid level.
For this purpose, we use a sequence of triangulations, and we perform
the smoothing only on their active levels. To create the sequence of
triangulations, one can use the functions \texttt{MGTransferGlobalCoarseningTools::create\_geometric\_coarsening\_sequence()}. A new version takes an
instance of \texttt{RepartitioningPolicyTools::Base} (see Subsection~\ref{sec:repartitioning}) as argument, which allows specifying the parallel
distribution of each multigrid level (in contrast to the fixed first-child policy in the case of local smoothing). These features have been developed and tuned for running on a supercomputer scale
with complicated coarse meshes as presented in \cite{kronbichler2021next}.
Furthermore, we added support for block vectors,
fixed a number of limitations, and performed performance optimizations of
the transfer operator; particularly, the redundant copy from/to temporary vectors
has been eliminated. Furthermore, hanging-node constraints are applied
efficiently in the same way as in the matrix-free loops (see Subsection~\ref{sec:mf}).

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \dealii{} was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order to judge the benefits
of one approach against the other, \dealii{} provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()}
in the \texttt{MGTools} namespace for the  estimation of the imbalance during, e.g.,
smoothing or the
communication efficiency during intergrid transfer, purely based on the given mesh.
% MK: I would not add this part, it does not really fit into this paper as
% there is nothing to report at this point (and we should then add references
% to the actual literature).
%The publication \cite{munch2022gc} also points out that not
%all types of smoothers are applicable for global coarsening due to the
%presence of hanging nodes, which is a motivation to add new smoother types
%to \dealii in the future.

@@

Table~\ref{tab:particle_timing} shows that all particle operations are
much faster in \dealii 9.4 than in version 9.3. In particular, operations that depend strongly on particle storage structure and require few fixed computations (like iteration and sorting) benefit massively from the above-mentioned optimizations. We note that the exact gains will depend strongly on the exact combination of geometry, mapping, dimensionality, and the number of particles per cell in any specific model, and can be smaller or larger than the measurements provided here.

@@

Many problems in parallel computations can be stated in the following
way: Each process in a parallel universe has a number of queries to
send to other processes that do not know that they will be asked, and
who will then have to respond with replies. This problem is solved by
``consensus algorithms''~\cite{hoefler2010scalable}. An example of where this problem appears is
given in Section~\ref{sec:repartitioning}.

@@

There are various ways to reference \dealii{}. To acknowledge the use of
the current version of the library, \textbf{please reference the present
  document}. For up-to-date information and a bibtex entry
see
\begin{center}
  \url{https://www.dealii.org/publications.html}
\end{center}
